Data-driven approaches to identifying knowledge gaps in marine invertebrate biodiversity

### 2017-05-28 manuscript.Rmd
update manuscript

---
#author: Fran\c{c}ois Michonneau
title: "Data-driven approaches to identifying knowledge gaps in marine invertebrate biodiversity"
output:
  pdf_document #:
    #pandoc_args: ["-F","pandoc-crossref"]
---

@@

<!---
the content of insect traps, marine sediments, plankton
samples, or communities settling on artificial substratum. When applied to
metazoan communities, these sequencing approaches typically target the barcoding
gene and therefore has been dubbed "meta-barcoding". These applications are also
making the public databases for these DNA barcodes all the more valuable. To
deliver on the meta-barcoding promises, it is crucial to have comprehensive
taxonomic coverage represented in the DNA sequences public databases. Here, we
investigate the proportion of species for each marine phylum that have been
sequenced at various spatial scale.
--->

@@

To compare the faunistic lists from the Gulf of Mexico and the Pacific Northwest with records in public databases, we queried the iDigBio and OBIS databases for all records they included in the same geographic boundaries as the lists. After filtering these lists to only include the same higher taxonomic level taxa as the species lists, we had for the Gulf of Mexico `r nrow(idigbio_gom_records)` and `r nrow(obis_gom_records)` records for iDigBio and OBIS respectively representing `r dplyr::n_distinct(idigbio_gom_records$worms_valid_name)` and `r dplyr::n_distinct(obis_gom_records$worms_valid_name)` species. For the Gulf of Mexico, most of the species included in the expert lists are recorded in both iDigBio and OBIS (xx %), or OBIS (xx %). However, for the Pacific Northwest, only xx% of the species from the expert list have records in public databases (Fig. 1).

@@

We extracted `r sum(gom_worms$is_marine, na.rm = TRUE)` marine invertebrates species from the faunistic list of the Gulf of Mexico, and `r sum(kozloff_worms$is_marine, na.rm=TRUE)` species from the Salish Sea. We then looked for the number of species listed that are known to occur in the same geographic area, the United States, or anywhere else in the world. Overall, OBIS had records for most of the species that are included in the species lists. However, even for the better studied phyla about 75% of the species have records from the same geographic area, while it is less than 50% for the Pacific Northwest.

These results suggest that even when considering species lists that have been curated and that represent possibly the most accurate knowledge for this geographic area, it would not be possible to reconstruct these lists based on currently digitized information. It highlights that additional digitization but even more importantly additional collecting is needed to validate the occurences of the species included in these lists.

>>>>

---
#author: Fran\c{c}ois Michonneau
title: "Data-driven approaches to identifying knowledge gaps in marine invertebrate biodiversity"
date: '`r format(Sys.time())`'
output:
  bookdown::pdf_document2:
    #includes:
    #  before_body: ../template/doc_prefix.tex
    #  in_header: ../template/preamble.tex
    keep_tex: yes
    latex_engine: pdflatex
    number_sections: no
    toc: no
  bookdown::word_document2: null
fontsize: 12pt
link-citations: yes
csl: nature.csl
bibliography: /home/francois/Library/dark-plankton.bib
---

@@

<!--
the content of insect traps, marine sediments, plankton
samples, or communities settling on artificial substratum. When applied to
metazoan communities, these sequencing approaches typically target the barcoding
gene and therefore has been dubbed "meta-barcoding". These applications are also
making the public databases for these DNA barcodes all the more valuable. To
deliver on the meta-barcoding promises, it is crucial to have comprehensive
taxonomic coverage represented in the DNA sequences public databases. Here, we
investigate the proportion of species for each marine phylum that have been
sequenced at various spatial scale.
-->

@@

To compare the faunistic lists from the Gulf of Mexico and the Pacific Northwest with records in public databases, we queried the iDigBio and OBIS databases for all records they included in the same geographic boundaries as the lists. After filtering these lists to only include the same higher taxonomic level taxa as the species lists, and keeping only records for which the species could be match to a species name in WoRMS, we had for the Gulf of Mexico `r nrow(idigbio_gom_records)` and `r nrow(obis_gom_records)` records for iDigBio and OBIS respectively representing `r dplyr::n_distinct(idigbio_gom_records$worms_valid_name)` and `r dplyr::n_distinct(obis_gom_records$worms_valid_name)` species. For the Pacific Northwest, we retrieved `r nrow(idigbio_kozloff_records)` and `r nrow(obis_kozloff_records)` records for iDigBio and OBIS respectively representing `r dplyr::n_distinct(idigbio_kozloff_records$worms_valid_name)` and `r dplyr::n_distinct(obis_kozloff_records$worms_valid_name)` species. For the Gulf of Mexico, the majority of the species included in the expert lists are recorded in both iDigBio and OBIS (`r format_output(species_in_common$n_spp_gom_idigbio_obis/species_in_common$n_spp_gom*100)` %), or OBIS (`r format_output(species_in_common$n_spp_gom_obis/species_in_common$n_spp_gom*100)` %) (Fig. \@ref(fig:database-overlap)). However, for the Pacific Northwest, only `r format_output(species_in_common$n_spp_kozloff_idigbio_obis/species_in_common$n_spp_kozloff*100)` and `r format_output(species_in_common$n_spp_kozloff_obis/species_in_common$n_spp_kozloff*100)` of the species from the regional list had records in OBIS and iDigBio or OBIS respectively (Fig. 1).

```{r compare_database_overlap, fig.cap='Overlap between the regional species list (list), records in iDigBio (idigbio), records in OBIS (obis), and records in BOLD (bold) for the Gulf of Mexico (left) and the Pacific Northwest (right). (\\#fig:database-overlap)', echo=FALSE}
knitr::include_graphics("figures/database_coverage.pdf")
```

@@

We extracted `r sum(gom_worms$is_marine, na.rm = TRUE)` marine invertebrates species from the faunistic list of the Gulf of Mexico, and `r sum(kozloff_worms$is_marine, na.rm=TRUE)` species from the Salish Sea. We then looked for the number of species listed that are known to occur in the same geographic area, the United States, or anywhere else in the world. Overall, OBIS had georeferenced records for most of the species that are included in the species lists. However, even for the better studied phyla about 75% of the species have records from the same geographic area, while it is less than 50% for the Pacific Northwest.

These results suggest that even when considering species lists that have been curated and that represent possibly the most accurate knowledge for this geographic area, today we could only reconstruct these list at 50 to 70% completedness based on digitized information. It highlights that additional digitization is necessary but even more importantly additional collecting is needed to validate the occurences of the species included in these lists.
